Oregon 2013 Medication for Asthma Inhalers

### 2017-07-18 r_scripts/analysis/oregon_medication.Rmd
Update oregon_medication.Rmd

```{r library, include=FALSE, echo=FALSE}
library(tidyverse)
# library(data.table)
library(survival)
library(ggplot2)
library(htmlTable)
library(knitr)
library(broom)
library(lme4)
library(lubridate) # date

@@

## Overview
For this research I have the Oregon All Payer All Claims (APAC) data set in 2013. I want to research on the effects of wildfire smoke on pharmacy outcome of asthma. NDC (National Drug Code) is the variable in the data set that the data set offered. The standard NDC I use the link below, which is a ndc list in 2013, which fits my original data set. 
(http://ww2.ncqa.org/hedis-quality-measurement/hedis-measures/hedis-2013/hedis-2013-final-ndc-lists)

@@

In these comparisons, we examine GWR of smoke PM2.5 estimations and associations with health outcomes using a time-stratified case-crossover study design. Pharmacy outcomes for a patient with a primary diagnosis of cardiopulmonary health outcomes and their date of admission (index time) were identified. We then created counterfactual observations for each patient on the same day of the week for the entire wildfire season (May 1st to Sep 30th, 2013). For patients who have a index time before May 1, 2013 or after September 30, 2013, their referent observations before or after these dates will be excluded as I will not be able to assign estimates of PM2.5 to these referent observations.

@@

There are about 77 million claims in the total data set. Some have the NDC but others are null or missing because people are going for pharmacy or seeing the doctor. So I choose the claims in Oregon State with ndc existing, which is about 23 million claims. Also, the POS (Place of service code) all show "pharmacy" type, comparing to the previous health disease outcomes we limit the "emergency visit" type or "urgent care" type.

For ndc, I choose the people who use the beta 2 agonists (one inhaler) for the day's first visit to form the sample. Because the ndc csv file has more information than what I need, I filter the ndc file by SQLite Studio (code and steps not attached) to a file only containing the beta 2 agonists medication information.

@@

For our research question, we focus on the GWR method (geographically weighted ridge regression). There are also WRF-Chem model (which substracts WRF-chem from the Weather Research and Forecasting with Chemistry no fire emission). For the WRF-Chem variable with the 'smoke' designator, this is WRF-Chem - WRF-Chem no fire. For Geo-Weighted Regression and Kriging model, with 'smk' designator, I subtracted off the 'Background' estimates of smoke, which I believe are the monthly averages of PM2.5 for a given grid. However, the 'Background' is the mean of total period from May 1st to Sep 30th 2013 so I have to assign the mean for every day. To reduce the occurance error and compare the differences I will analyze all three models.

### Make category index for beta-2 agonists.
First I filtered the whole data set with ndc existing and only Oregon state. Then there are about 24 million claims left. Then I filtered the data set with only beta-2-agonists and found about 233 thousand claims (83787 unique person). By using these unique personkey, I found the claims of them in the total data set, which is about 8.9 million claims. This the data set that I use for the category classification.

@@

@@

I also make time stratified time series data set. The original data set only contains the zip code, but for time series, we want to research on the county level. Therefore, I assign each zip code with one county which I found online. There are 488 zipcodes in total (4 are missing but I use google to find it).

@@

plot_ts <- ggplot(or_ndc_county_ts, aes(y =rate_per_100k, x = date)) +
           geom_jitter()

@@

plot_county <- ggplot(or_ndc_county_ts, aes(y =rate_per_100k, x = date)) +
               geom_jitter() + facet_wrap(~county)

@@

### check if the time series has the association with get smoke
a <- glm(rate_per_100k ~ geo_smk_pm10, data = or_ndc_county_ts, family = "poisson")
summary(a)

plot_health_smk <- ggplot(or_ndc_county_ts, aes(x = geo_smk_pm10, y = log(rate_per_100k+0.1))) +
  geom_point(aes(colour = c("red")), alpha = 0.2,  size = 1) + 
  # scale_colour_manual(values = c("red", "blue")) +
  # truncated y axis
  #scale_y_continuous(limits = c(0, 1.5)) +
  # facet_wrap(~county, scale  = "free") +
  ylab("Log Outcome Rate per 100,000 Persons") +
  xlab("GWR Smoke PM2.5 per 10*10ug/m^3") +
  theme_bw()

# unnecessary
plot_county_health_smk <- ggplot(or_ndc_county_ts, aes(x = geo_smk_pm10, y = log(rate_per_100k+0.1))) +
  geom_point(aes(colour = county), alpha = 0.2,  size = 1) + 
  # scale_colour_manual(values = c("red", "blue")) +
  # truncated y axis
  #scale_y_continuous(limits = c(0, 1.5)) +
  facet_wrap(~county, scale  = "free") +
  ylab("Log Outcome Rate per 100,000 Persons") +
  xlab("GWR Smoke PM2.5 per 10*10ug/m^3") +
  theme_bw()

@@

ggsave("../plot_new/ndc_county_rate_health_smk.pdf", plot = plot_county_health_smk, 
       width = 12, height = 8, units = "in")

@@

For all of 2013, there were a total of millons of all claims records in the APAC dataset. During the May 1st to September 30th wildfire season in Oregon state, there were 24217 pharmacy visits for the ndc of beta 2 agonists. The number of cases, as well as age-specific and sex-specific strate are shown below.

@@

# dataframe list
method_list <- c('WRF-Chem Smoke', 'Geo-Weighted Smoke', 'Kriging Smoke')

# outcome name
outcome_name <- "Inhalers"

@@

tab <- htmlTable(txtRound(asthma_point_est_df, digits = 1), 
           caption = "Number of cases for each outcome observed from May 1st to September 30st, 2013",
           # column headers
           header = c("Medication", "Cases n", "Less than 15", "(%)", "15 to 65", "(%)", "Greater than 65",
                      "(%)", "Female", "(%)", "Male", "(%)"),
           # column spanner
           cgroup = c("","Age Category", "Sex"), 
           n.cgroup = c(2, 6, 4),
           padding.rgroup = "&nbsp;&nbsp;",
           css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
           align = "llccccc" # column alignment,
            ) # end table

@@

### Overall plot

@@

# outcome name
outcome <- "Asthma Inhalers"
outcome_name <- "Asthma Inhalers"

@@

  print(print_plot)
  # save figure
  ggsave("../plot_new/ndc_overall_three.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@

For some outcomes, such like Age > 65 in WRF-Chem method, the CI is the only insignificant one in the plot.

@@

All the CI do not include 1, which means the use of asthma medication (beta 2 agonists) increases about 5% when smoke increase 10 units. So the risk for getting asthma will increase about 5% ~ 8% when the smoke PM2.5 increase 10 units.

>>>>

```{r library, include=FALSE, echo=FALSE}
library(tidyverse)
# library(data.table)
library(survival)
library(ggplot2)
library(htmlTable)
library(knitr)
library(broom)
library(lme4)
library(maps)
library(lubridate) # date

@@

## Overview

We had previously shown GWR to be a good estimate of wildfire smoke concentrations, and showed associations with respiratory outcomes (including asthma) in Washington state. We want to replicate results for Oregon and also look at beta2 agonist pharmacy fills.

For this research I have the Oregon All Payer All Claims (APAC) data set in 2013. I want to research on the effects of wildfire smoke on beta-2 agonist pharmacy fill, a medication primarily used as a rescue medication for persons with asthma or COPD. NDC (National Drug Code) is the variable in the data set that the data set offered. The standard NDC I use the link below, which is a ndc list in 2013, which fits my original data set. 
(http://ww2.ncqa.org/hedis-quality-measurement/hedis-measures/hedis-2013/hedis-2013-final-ndc-lists)

@@

In these comparisons, we examine GWR method of smoke PM2.5 estimations and associations with health outcomes using a time-stratified case-crossover study design. Pharmacy outcomes for a patient with a primary diagnosis of all pharmacy records from the state, and it just so happens we also have APAC data for people that were billed for a cardiopulmonary event that we can link to the pharmacy claim, and their date of admission (index time) were identified. We then created counterfactual observations for each patient on the same day of the week for the entire wildfire season (May 1st to Sep 30th, 2013). For patients who have a index time before May 1, 2013 or after September 30, 2013, their referent observations before or after these dates will be excluded as I will not be able to assign estimates of PM2.5 to these referent observations.

@@

There are about 77 million claims in the total data set, of which 23 million are pharmacy records denoted by the NDC variable. So I choose the claims in Oregon State with ndc existing, which is about 23 million claims. Also, the POS (Place of service code) all show "pharmacy" type, comparing to the previous health disease outcomes we limit the "emergency visit" type or "urgent care" type.

For ndc, I choose the people the person was billed for filling a beta2 agonist prescription, which we use as a proxy for use to form the sample. Because the ndc csv file has more information than what I need, I filter the ndc file by SQLite Studio (code and steps not attached) to a file only containing the beta 2 agonists medication information.

@@

@@

For our research question, we focus on the GWR method (geographically weighted ridge regression). For Geo-Weighted Regression model, with 'smk' designator, I subtracted off the 'Background' estimates of smoke, which I believe are the monthly averages of PM2.5 for a given grid. However, the 'Background' is the mean of total period from May 1st to Sep 30th 2013 so I have to assign the mean for every day.

@@

I also make time stratified time series data set. The original data set only contains the zip code, but for time series, we want to research on the county level. Therefore, I assign each zip code with one county which I found online. There are 488 zipcodes in total (4 are missing but I use google to find it). This is a sensitivity analysis to confirm the casecross over results.

@@

plot_ts <- ggplot(or_ndc_county_ts, aes(y =rate_per_100k, x = date)) +
           geom_jitter() +
           ylab("Rate of 100k of Beta2 agonist fills") +
           xlab("Year 2013")

@@

plot_county <- ggplot(or_ndc_county_ts, aes(y =rate_per_100k, x = date)) +
               geom_jitter() + facet_wrap(~county) +
               ylab("Rate of 100k of Beta2 agonist fills")

@@

plot_smk_ts <- ggplot(or_ndc_county_ts, aes(y =geo_smk_pm10, x = date)) +
           geom_jitter() +
           ylab("GWR smoke concentration (10ug/m^3)") +
           xlab("Year 2013")

print(plot_smk_ts)

### check if the time series has the association with get smoke
a <- glm(rate_per_100k ~ geo_smk_pm10, data = or_ndc_county_ts, family = "poisson")
# summary(a)

plot_health_smk <- ggplot(or_ndc_county_ts, aes(x = geo_smk_pm10, y = log(rate_per_100k+0.1))) +
  geom_point() + 
  # geom_point(aes(colour = c("red")), alpha = 0.2,  size = 1) + 
  # scale_colour_manual(values = c("red", "blue")) +
  # truncated y axis
  #scale_y_continuous(limits = c(0, 1.5)) +
  # facet_wrap(~county, scale  = "free") +
  ylab("Log Beta 2 agonists Rate per 100,000 Persons") +
  xlab("GWR Smoke PM2.5 per 10ug/m^3") +
  theme_bw()

# unnecessary
# plot_county_health_smk <- ggplot(or_ndc_county_ts, aes(x = geo_smk_pm10, y = log(rate_per_100k+0.1))) +
#   geom_point(aes(colour = county), alpha = 0.2,  size = 1) + 
  # scale_colour_manual(values = c("red", "blue")) +
  # truncated y axis
  #scale_y_continuous(limits = c(0, 1.5)) +
#   facet_wrap(~county, scale  = "free") +
#   ylab("Log Outcome Rate per 100,000 Persons") +
#   xlab("GWR Smoke PM2.5 per 10*10ug/m^3") +
#   theme_bw()

@@

# ggsave("../plot_new/ndc_county_rate_health_smk.pdf", plot = plot_county_health_smk, 
#        width = 12, height = 8, units = "in")

### map of smoke days

```{r}
meta_ts <- read_csv("../data/meta_timeseries.csv") %>% 
  # creating 10 unit increase var of geo smoke and day of week
  mutate(geo_smk10 = geo_smk_pm/10,
         day = weekdays(date)) %>% 
  # create smoke and lag variables
  mutate(smoke0 = ifelse(geo_smk_pm > 0, 1, 0),
         smoke5 = ifelse(geo_smk_pm > 5, 1, 0),
         smoke10 = ifelse(geo_smk_pm > 10, 1, 0),
         smoke15 = ifelse(geo_smk_pm > 15, 1, 0), 
         # lag variable
         geo_smk_lag1 = lag(geo_smk10, 1, order_by = county),
         geo_smk_lag2 = lag(geo_smk10, 2, order_by = county),
         geo_smk_lag3 = lag(geo_smk10, 3, order_by = county),
         geo_smk_lag4 = lag(geo_smk10, 4, order_by = county),
         geo_smk_lag5 = lag(geo_smk10, 5, order_by = county),
         geo_smk_lag6 = lag(geo_smk10, 6, order_by = county),
         geo_smk_lag7 = lag(geo_smk10, 7, order_by = county)) %>% 
  ungroup()

# smoke wave days (at least 2 consecutive days in the 98th percentile of PM)
quantile(meta_ts$geo_wt_pm, probs = c(0.95,0.98,0.99, 0.999))
# following Coco's smoke wave day, the 98th percentile for PM2.5 concentrations
# would be >26.9

smoke_wave <- meta_ts %>% 
  select(county, date, geo_wt_pm) %>% 
  mutate(high_pm_day = ifelse(geo_wt_pm > 26.9, 1, 0)) %>% 
  group_by(county) %>% 
  # identify consecutive days
  mutate(smoke_wave = ifelse(lag(high_pm_day, 
    order_by = county)==1 & high_pm_day ==1, 1, 0)) %>% 
  select(-high_pm_day, -geo_wt_pm)

# merge in smokewave day
meta_ts <- meta_ts %>% 
  right_join(smoke_wave, by = c("county", "date"))

# count of state outcome timeseries
state_outcome_rate <- meta_ts %>% 
  group_by(date) %>% 
  # remove columns I don't want to calculate
  select(-fips, -county) %>% 
  # add up outcomes in each county
  summarise_each(funs(sum), population, n_obs:broken_arm_n) %>% 
  mutate_each(funs((./population*100000)), n_obs:broken_arm_n)

# average of environmental measures, joined with outcomes
state_smoke <- meta_ts %>% 
  group_by(state, date) %>% 
  select(-fips, -county) %>% 
  summarise_each(funs(mean, median, min, max), wrf_pm:wrf_temp) %>% 
  full_join(state_outcome_rate, by = c("date"))

# only 150 smoke wave days
xtabs(~smoke_wave,  meta_ts)
# if we compare to our geo smoke > 10, we'd add 362 extra days
xtabs(~smoke_wave + smoke10, meta_ts)
# if we used smoke >15 we'd add 164 extra days 
xtabs(~smoke_wave + smoke15, meta_ts)

# small multiples dataset
or_smk <- or_ts %>% 
  select(county, date, wrf_smk_pm, krig_smk_pm, geo_smk_pm) %>% 
  gather(key = smk_method, value = pm, -date, -county)

or_smk_plot <- ggplot(or_smk, aes(x = date, y = pm)) +
  geom_point(aes(colour = smk_method), alpha = 0.8, size = 0.8) +
  facet_wrap(~county) +
  ggtitle("Oregon: 2013 Fire Season Smoke PM2.5") +
  ylab("Smoke PM2.5 ug/m^3") +
  xlab("Year: 2013") +
  theme_bw()

or_smk_plot

# data wrangle ----
#first step is to summarise the counts of smoke days in each county 
or_smk_count <- meta_ts %>% 
  group_by(county) %>% 
  # summing up smoke days
  summarise(smk_wave_n = sum(smoke_wave), smk_day0 = sum(smoke0), 
            smk_day5 = sum(smoke5), smk_day10 = sum(smoke10),
            smk_day15 = sum(smoke15)) %>% 
  # county needs to be lowercase to join to maps dataframe
  mutate(county = tolower(county))

# maps package will be loaded to pull map data and create a dataframe
or_county_df <- map_data("county", "oregon")

# join county counts to map df
or_smk_df <- or_county_df %>% 
  right_join(or_smk_count, by = c("subregion" = "county")) %>% 
  # small multiples of different smoke cutoff methods
  gather(key = smk_cut_method, value = smoke_days, -long:-subregion) %>% 
  # filter out smoke day 0 since it's not very specific
  filter(smk_cut_method != "smk_day0") %>% 
  # rename smoke day cutoffs 
  mutate(smk_cut_method = 
    ifelse(smk_cut_method == "smk_day5","Smoke PM2.5 > 5 ug/m^3",
    ifelse(smk_cut_method == "smk_day10", "Smoke PM2.5 > 10 ug/m^3",
    ifelse(smk_cut_method == "smk_day15", "Smoke PM2.5 > 15 ug/m^3",
    ifelse(smk_cut_method == "smk_wave_n", "Smoke Wave", NA)))),
    # preserve smoke method order for small multiple
    smk_cut_method = factor(smk_cut_method, 
      levels = c("Smoke PM2.5 > 5 ug/m^3", "Smoke PM2.5 > 10 ug/m^3", 
                 "Smoke PM2.5 > 15 ug/m^3", "Smoke Wave")))

# map ----

# smoke days where GWR smoke > 15
smoke_map <- ggplot(or_smk_df, aes(x=long, y=lat, group=group)) +
  # fill with number of smoke days
  geom_polygon(aes(fill = smoke_days), alpha = 0.7) +
  scale_fill_gradient("Smoke-Impacted Days",
                      low = "#0b486b", high = "#f56217") +
  # add county path on top
  geom_path() +
  facet_wrap(~smk_cut_method) +
  xlab("Longitude") +
  ylab("Latitude") +
  ggtitle("Oregon: Counties affected by smoke") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(colour=NA, fill=NA))

smoke_map

@@

For all of 2013, there were a total of 77 millon claim records in the APAC dataset. During the May 1st to September 30th wildfire season in Oregon state, there were 24217 pharmacy visits for the ndc of beta 2 agonists. The number of cases, as well as age-specific and sex-specific strate are shown below.

@@

# outcome name
outcome_name <- "beta-2 agonist"

@@

tab <- htmlTable(txtRound(asthma_point_est_df, digits = 1), 
           caption = "Total number of cases for pharmacy fills of beta-2 agonist observed from May 1st to September 30st, 2013",
           # column headers
           header = c("Medication", "Cases n", "Less than 15", "(%)", "15 to 65", "(%)", "Greater than 65",
                      "(%)", "Female", "(%)", "Male", "(%)"),
           # column spanner
           cgroup = c("","Age Category", "Sex"), 
           n.cgroup = c(2, 6, 4),
           padding.rgroup = "&nbsp;&nbsp;",
           css.cell = "padding-left: 0.5em; padding-right: .5em;", # cell space
           align = "llccccc" # column alignment,
            ) # end table

@@

### Overall plot with GWR method

@@

# outcome name
outcome <- "Beta-2 agonists"
outcome_name <- "Beta-2 agonists"

@@

  print(print_plot)
  # save figure
  ggsave("../plot_new/ndc_overall_gwr.pdf", plot = print_plot, 
       width = 12, height = 8, units = "in")

@@

For some outcomes, such like Age > 65 in WRF-Chem method, the CI is the only not significant one in the plot.

The plot shows that overall, the risk of pharmacy fills of beta-2 agonist will increase significantly as the smoke concentrate increases. For age group of 15-65, the risk is the highest, for 10 ug/m^3 increase of pm2.5, the risk of filling the beta-2 agonist will increase about 9%. For age groups which are less than 15 or more than 65, the risk is similar, both of them will increase about 5.6%.

@@

All the CI do not include 1, which means the pharmacy filling (beta 2 agonists) increases about 5% when smoke increase 10 units. So the risk for getting asthma will increase about 5% ~ 8% when the smoke PM2.5 increase 10 units. Also, the risk for female is much higher than that for male.

These results show that there may be effect modification by sex, where females might be more likely to fill a beta-2 agonist on the same day it's really smoky for the GWR estimates.
